In this section, we establish our results regarding what we call ``cyclic'' and ``unary'' simple functions. First, we formally define these notions.

\begin{definition}[Cyclic simple functions]
    Given a root-free $v \in \Gamma^*$, we say that $f : A \xrightarrow{\mathrm{sf}} \Gamma^*$ is $v$-cyclic if
    \[
        \im (f) \subseteq v^*.
    \]
\end{definition}

\begin{definition}[Unary simple functions]
    Given a root-free $v \in \Gamma^*$, we say that $f : A \xrightarrow{\mathrm{sf}} \Gamma^*$ is $v$-unary if
    \[
        \im (f) \subseteq \infix \left( v^\omega \right).
    \]
\end{definition}

Clearly, as stated, the notion of unarity is weaker than cyclicity. However, in the following part of this chapter, we will show that, as long as simple functions are concerned, being unary is almost the same as being cyclic. This is great news for us, because it is therefore sufficient for us to develop a system for proving cyclicity of simple functions, as basically all reasonable notions of ``unarity'' fall into this category. 

In the following part of this section, we introduce several new axioms and lemmas to our equational system, which will be used for proving cyclicity. Then, we prove that unarity is essentially the same as cyclicity in the world of simple functions. Finally, at the end of this chapter, we prove that our system is complete when it comes to statements of the form ``$f$ is $v$-cyclic''.

\subsubsection{Axioms and lemmas for cyclicity}

Before we proceed any further, we have to address the elephant in the room: the statements of the form ``$f$ is cyclic'' are not equational assertions! Fortunatelly, this one is pretty simple to overcome. We claim that

\begin{center}
    $f$ is $v$-cyclic \quad if and only if \quad $f^{|v|} = v^{|f|}$. 
\end{center}

Note that both $f^{|v|}$ and $v^{|f|}$ are simple functions: the former is $f$ concatenated a finite number of times and the latter is $f$ post-composed with a homomorphism.

In fact, we show a more general property for simple functions on $\Gamma^{(*)}$. Given a function $f : A \xrightarrow{\mathrm{sf}} \Gamma^{(*)}$, by $v^f$ we denote $f$ post-composed with the homomorphism which maps $\Gamma$ to $v$ and $\overline{\Gamma}$ to $\overline{v}$. Clearly, $v^f$ is a simple function. With this notion in hand, we introduce an equational definition of cyclicity of simple functions on the free group.

\begin{claim}
    If $v \in \Gamma^*$ is root-free, then $\im(f) \subseteq v^{(*)}$ if and only if $f^{|v|} = v^f$. 
\end{claim}

\begin{proof}[Proof of $\implies$]
    Let $a \in \dom(f)$ be an arbitrary input. If $f(a) = v^\ell$ for some $\ell \geq 0$, then $f(a)^{|v|} = v^{\ell |v|} = v^{|f(a)|} = v^f$. Similarly, if $f(a) = \overline{v}^\ell$ for some $\ell \geq 0$, then $f(a)^{|v|} = \overline{v}^{\ell |v|} = \overline{v}^{|f(a)|} = v^f$.
\end{proof}

\begin{proof}[Proof of $\Longleftarrow$]
    Let $a \in \dom(f)$ be an arbitrary input. We know that $f(a)^{|v|} = v^f = v^\ell$ for some $\ell \in \mathbb{Z}$. If $\ell = 0$, then $f(a) = \varepsilon \in v^{(*)}$. However, if $\ell \neq 0$, then by \cref{lem:common-root-theorem} we obtain that $f(a)$ and $v$ have a common root. Since $v$ is root-free, we deduce that $f(a)$ is a power of $v$.
\end{proof}

In light of the equational definition of cyclicity, whenever we use $\im(f) \subseteq v^*$, we can treat it as syntax sugar for $f^{|v|} = v^{|f|}$. Similarly, we treat $\im(f) \subseteq v^{(*)}$ as $f^{|v|} = v^f$. We are now in position to introduce all new axioms needed for proving cyclicity: 

\[
    \im(\varepsilon) \subseteq v^{(*)} \quad \im(v) \subseteq v^{(*)} \quad \im(\overline{v}) \subseteq v^{(*)} 
\]

\begin{mathpar}
    \inferrule
        {
            \im(f) \subseteq v^{(*)} \\ 
            \im(g) \subseteq v^{(*)} 
        }
        {
            \im(f \cdot g) \subseteq v^{(*)}
        }[\textsc{(unarity-concat-lift)}]
\end{mathpar}

\begin{mathpar}
    \inferrule
        {   
            \im (f) \subseteq v^{(*)} \\ 
            \im (g) \subseteq v^{(*)} 
        }
        {
            f \cdot g = g \cdot f
        }[\textsc{(unarity-comm)}]\label{rule:unarity-commutativity}
\end{mathpar}

\begin{mathpar}
    \inferrule
        {   
            \im (h) \subseteq v^{(*)}
        }
        {
            \im \left( \bihom{to, h} \right) \subseteq v^{(*)}
        }[\textsc{(unarity-bihom-lift)}]\label{rule:unarity-bihom-lift}
\end{mathpar}

\begin{mathpar}
    \inferrule
        {   
            \im (g) \subseteq v^{(*)} \\ 
            \im (h) \subseteq v^{(*)}
        }
        {
            \bihom{to, g \cdot h} = \bihom{to, g} \cdot \bihom{to, h}
        }[\textsc{(unarity-bihom-dist)}]\label{rule:unarity-bihom-dist}
\end{mathpar}

Apart from the axioms, we give two lemmas we will use to prove cyclicity by structural induction. One is for concatenation of simple functions, and the other is for bihomomorphisms:

\begin{mathpar}
    \inferrule
        {
            {\red{f, g : A \to \Gamma^* \quad x,y \in \Gamma^*}} \\\\
            \im \left(f \cdot x \right) \subseteq v^* \\ 
            \im \left(y \cdot g \right) \subseteq v^* \\ 
            y \cdot x = v
        }
        {
            \im \left(f \cdot g \right) \subseteq v^*
        }[\textsc{(unarity-concat)}]\label{rule:unarity-concat}
\end{mathpar}

\begin{mathpar}
        \inferrule
        {
            {\red{h_{| A_i \times A_j} : A_i \times A_j \to \Gamma^* \quad u_i, w_i \in \Gamma^*}} \\\\
            \forall i,j \quad \im \left( u_i \cdot h_{| A_i \times A_j} \cdot w_j \right) \subseteq v^* \\
            \forall i \quad u_i \cdot w_i = v
        }
        {
            \im \left( u_1 \cdot \bihom{to, h} \cdot w_1 \right) \subseteq v^*
        }[\textsc{(unarity-bihom)}]\label{rule:unarity-bihom}
\end{mathpar}

\katzpermargin{Are the red assumptions in the lemmas \nameref{rule:unarity-concat} and \nameref{rule:unarity-bihom} legal? Can we have separate rules for functions with monoid outputs? Or should we treat "monoid" functions as a syntactic property?}

We prove soundness of \nameref{rule:unarity-concat} and \nameref{rule:unarity-bihom} in two ways. In this section, we provide more relaxed arguments, in order to build some intuition behind the lemmas. In \cref{sec:proofs-of-lemmas}, we give formal derivations from the axioms.

\begin{claim}
    The lemma \nameref{rule:unarity-concat} is sound.
\end{claim}

\begin{proof}
    For every input $a$, we can rewrite $f(a) \cdot g(a)$ as follows:
    \[
        f(a) \cdot g(a) = f(a) \cdot x \cdot \overline x \overline y \cdot y \cdot g(a) \in v^* \cdot \overline{v} \cdot v^* \in v^{(*)}.
    \]
    Since $\im(f), \im(g) \subseteq \Gamma^*$, we deduce that $f(a) \cdot g(a) \in v^*$. 
\end{proof}

\begin{claim}
    The lemma \nameref{rule:unarity-bihom} is sound.
\end{claim}

\begin{proof}
    Let $a = [a_1, \dots, a_n]$ be an arbitrary input for $\bihom{to, h}$. Let $t_i := \type{a_i}$ for every $i \in [n]$. We write:
    \[
        \bihom{to, h}(a) = \prod_{i=1}^{n-1} h(a_i, a_{i+1}) \in \prod_{i=1}^{n-1} \overline{u_{t_i}} \cdot v^* \cdot \overline{w_{t_{i+1}}} \in \overline{u_1} \cdot v^{(*)} \cdot \overline{w_1}.
    \]
    This gives $u_1 \cdot \bihom{to, h}(a) \cdot w_1 \in v^{(*)}$. Since $\im(h) \subseteq \Gamma^*$, the proof is finished. 
\end{proof}

\subsubsection{Unary simple functions are basically cyclic}

Clearly, unarity is a weaker assumption than cyclicity. However, it turns out that, when it comes to simple functions with inputs in the free monoid, these two notions are, more or less, equivalent. 

\begin{lemma}\label{lem:unary-is-the-same-as-cyclic}
    Suppose that $v \in \Gamma^*$ is root-free and $f : A \xrightarrow{\mathrm{sf}} \Gamma^*$ is $v$-unary. Then there are $u,w \in \Gamma^*$ such that 
    \[
        \im(u \cdot f \cdot w) \subseteq v^*. 
    \]
\end{lemma}

The rest of this section is devoted to proving \cref{lem:unary-is-the-same-as-cyclic}. By proceed by a nested induction on the input type rank, and then the number of subfuctions. The statement clearly holds if the input is an atomic type, or more generally, $f$ is a constant function. We consider the cases of concatenation and bihomomorphism separately. However, before we proceed to that, we prove the following auxiliary lemma.

\begin{lemma}\label{lem:unarity-you-can-attach-prefix-to-unbounded}
    Suppose that $f,g : A \xrightarrow{\mathrm{sf}} \Gamma^*$, $u_0,u_1,w_0,w_1 \in \Gamma^*$, $f$ is unbounded, $(u_0 f w_0)$ and $(u_1 g w_1)$ are $v$-cyclic and $(f g)$ is $v$-unary. Then $(u_0 f g)$ is $v$-unary.
\end{lemma}

\begin{proof}
    By cyclicity, every output of $(u_0 f g)$ is of the form 
    \[
        u_0 \cdot (\overline{u_0} \cdot v^k \cdot \overline{w_0}) \cdot (\overline{u_1} \cdot v^\ell \cdot \overline{w_1})
    \]
    for some $k, \ell \geq 0$. Let $a : A$ be such that $(u_0 f g)(a)$ represented as above has $k \geq 3$ (which can be achieved since $f$ is unbounded), and $\ell \geq 3$ if possible, otherwise $g$ is constant and the choice of $a$ does not affect $\ell$. By \cref{lem:copies-of-v-can-be-pumped}, if we show that $(u_0 f g)(a) \leq_\infix v^\omega$, then we are done. Since $k \geq 3$, we have that 
    \[
        (f \cdot g)(a) \in (\overline{u_0} v) \cdot v^+ \cdot (v \overline{w_0}) \cdot (\overline{u_1} \cdot v^\ell \cdot \overline{w_1}).
    \]
    Therefore, by \cref{lem:copy-of-v-in-infix-splits-into-prefix-and-suffix} we know that $(v \overline{w_0} \cdot \overline{u_1} \cdot v^\ell \cdot \overline{w_1})$ is prefix of $v^\omega$, yielding $u_0 \cdot (fg)(a) \leq_\infix v^\omega$.
\end{proof}

\paragraph{Concatenation.} Suppose that $f g$ is $v$-unary. Then both $f$ and $g$ are $v$-unary, and by induction, there are $u_0,w_0,u_1,w_1 \in \Gamma^*$ such that $\im(u_0 f w_0), \im(u_1 g w_1) \subseteq v^*$. We can assume that $|u_0|, |u_1| \in \{0, \dots, |v|-1\}$ and $|w_0|, |w_1| \in \{1, \dots, |v| \}$. If both $f,g$ are constant, then so is $(f g)$ and we are done. Hence, we now assume that $(f g)$ is not constant. 

Next, suppose that exactly one of $f,g$ is constant. By symmetry, it suffices to consider the case where $g$ is constant and $f$ is unbounded. Suppose that $g = c$ for some $c \in \Gamma^*$. By \cref{lem:unarity-you-can-attach-prefix-to-unbounded}, we known that $(u_0 f c)$ is $v$-unary. Now the statement is a consequence of the following claim for $f := (u_0 f)$.

\begin{claim}\label{claim:cyclicity-concat-attaching-constant}
    Suppose that $f : A \xrightarrow{\mathrm{sf}} \Gamma^*$, $w_0, c \in \Gamma^*$, $f$ is unbounded, $(f w_0)$ is $v$-cyclic and $(f c)$ is $v$-unary. Then there is $w \in \Gamma^*$ such that $(f  c  w)$ is $v$-cyclic.
\end{claim}

\begin{proof}
    Let $a : A$ be such that $f(a) \cdot w_0 \in v^k$ for some $k \geq 2$. Then 
    \[
        f(a) \cdot c \in v^+ \cdot (v \overline{w_0}) \cdot c.
    \]
    By \cref{lem:copy-of-v-in-infix-splits-into-prefix-and-suffix}, we deduce that $v \overline{w_0} c$ is a prefix of $v^\omega$. Let $w \in \Gamma^*$ be such that $v \overline{w_0} c w \in v^*$. Then 
    \[
        \im(f c w) \subseteq v^* \cdot \overline{w_0} \cdot c \cdot w \subseteq v^{(*)} \cdot (v \overline{w_0} c w) \subseteq v^{(*)}.
    \]
    Since $f,c,w$ are all elements of the free monoid, the statement follows.
\end{proof}

The remaining case is that both $f,g$ are unbounded. By \cref{lem:unarity-you-can-attach-prefix-to-unbounded}, $(u_0 f \cdot g w_1)$ is $v$-unary. Observe that it suffices to show that $u_1 \cdot w_0 \in v^*$ --- by the choice of lengths of $u_1$ and $w_0$ this will give $u_1 \cdot w_0 = v$ and the statement will follow by \nameref{rule:unarity-concat}. 

Thus, the following claim, applied to $f := (u_0 f), g:= (g w_1)$ finishes the argument in the case of concatenation.

\begin{claim}\label{claim:cyclicity-concant-both-unbounded}
    Suppose that $f,g : A \xrightarrow{\mathrm{sf}} \Gamma^*$ are unbounded, $(f g)$ is $v$-unary and $(f x)$, $(y g)$ are $v$-cyclic for some $x,y \in \Gamma^*$. Then $(y x) \in v^*$. 
\end{claim}

\begin{proof}
    Take $a : A$ such that $f(a) \in v^+ \cdot (v^k \overline x)$ and $g(a) \in (\overline y v^\ell) \cdot v^+$, where $k, \ell \geq 0$ and $(v^k \overline x), (\overline y v^\ell) \in \Gamma^*$. Then we have 
    \[
        (f \cdot g)(a) \in v^+ \cdot (v^k \cdot \overline x)(\overline y \cdot v^\ell) \cdot v^+.
    \] 
    By \cref{lem:between-two-copies-of-v-it-must-be-power-of-v}, we get that $v^k \cdot \overline x \overline y \cdot v^\ell \in \Gamma^*$, and therefore $(y x) \in v^*$. 
\end{proof}

\paragraph{Bihomomorphisms.} Suppose that $f = \bihom{to, h} : A_1 \times (A_1 + \dots + A_m)^* \times A_1 \to \Gamma^*$ (the case of reversed bihomomorphism is symmetric). We can assume that $f \neq \varepsilon$, and therefore $f$ is unbounded. The function $h$ can be viewed as a collection of simple functions $h_{i,j} : A_i \times A_j \to \Gamma^*$. Clearly, $h_{i,j}$ is $v$-unary for every $i,j \in [m]$. First, we show that for every $i,j \in [m]$, how outputs of $h_{i,j}$ are ``cut'' at the beginning only depends on $i$, and how they are ``cut'' at the end only depends on $j$.

\begin{claim}\label{claim:cyclicity-bihom-prefix-and-suffix-depends-only-on-bound-type}
    There are $u_1, \dots, u_m, w_1, \dots, w_m \in \Gamma^*$ such that for every $i,j \in [m]$, we have $\im(u_i \cdot h_{i,j} \cdot w_i) \subseteq v^*$. 
\end{claim}

\begin{proof}
    First, we describe how to choose $u_1, \dots, u_m, w_1, \dots, w_m$, and then we prove that our choice satisfies the statement. 

    For every $i \in [m]$, we do the following. Pick a sequence $a^{(i)} = a_1, a_2, \dots, a_n$ such that (1) $a_1 : A_1$, (2) $a_n : A_i$, and (3) $h^*([a_1, \dots, a_{n-1}])$ is long enough to contain a copy of $v^3$. Consider the simple function $g_0 : \prod_{j=1}^n \type{a_j} \to \Gamma^*$ defined as 
    \[
        g_0(a'_1, \dots, a'_n) = h^*([a'_1, \dots, a'_n]).
    \]
    The function $g_0$ has input type of smaller rank and is $v$-unary, so from induction, there are $x_i, y_i \in \Gamma^*$ such that $\im(x_i g_0 y_i) \subseteq v^*$. We can assume that $|x_i|, |y_i| < |v|$. With this in hand, we let $u_i := v \overline{y_i}$.  

    The word $w_i$ is chosen in a similar way. Pick a sequence $b^{(i)} = b_1, b_2, \dots, b_k$ such that (1) $b_1 : A_i$, (2) $b_k : A_1$, and (3) $h^*([b_2, \dots, b_k])$ contains a copy of $v^3$. Then we consider $g_1 : \prod_{j=1}^k \type{b_j} \to \Gamma^*$ defined as $g_1(b'_1, \dots, b'_k) = h^*([b'_1, \dots, b'_k])$. By induction, there are $z_i, s_i \in \Gamma^*$ with $|z_i|, |s_i| < |v|$ such that $\im(z_i g_1 s_i) \subseteq v^*$. Now we let $w_i := \overline{z_i} v$.  

    We now prove that our choice satisfies the statement. Take any $i,j \in [m]$ and a tuple $(a_i, a_j) \in A_i \times A_j$. Let $a$ be equal to $a^{(i)}$, but with the last element changed to $a_i$. Similarly, let $b$ be the list $b^{(j)}$, but with the first element equal to $a_j$. Consider the following output:
    \begin{align*}
        f(a \cdot a_i \cdot a_j \cdot b) &= h^*(a) \cdot h_{i,j}(a_i, a_j) \cdot h^*(b) \\ 
                                         &= g_0(a) \cdot h_{i,j}(a_i, a_j) \cdot g_1(b) \\ 
                                         &\in \overline{x_i} \cdot v^{\geq 3} \cdot \overline{y_i} \cdot h_{i,j}(a_i, a_j) \cdot \overline{z_j} \cdot v^{\geq 3} \cdot \overline{s_j} \\ 
                                         &\in (\overline x_i v) \cdot v^+ \cdot (v \overline y_i) \cdot h_{i,j}(a_i, a_j) \cdot (\overline z_j v) \cdot v^+ \cdot (v \overline s_j).
    \end{align*}
    By \cref{lem:between-two-copies-of-v-it-must-be-power-of-v}, we get that $(v \overline{y_i}) \cdot h_{i,j}(a_i, a_j) \cdot (\overline{z_j} v) \in v^*$, which is exactly what we needed to prove.
\end{proof}

Let $u_1, \dots, u_m, w_1, \dots, w_m$ be as in \cref{claim:cyclicity-bihom-prefix-and-suffix-depends-only-on-bound-type}. We can assume that $|u_i| \in \{0, \dots, |v|-1 \}$ and $|w_i| \in \{1, \dots, |v| \}$ for every $i \in [m]$. By \nameref{rule:unarity-bihom}, it remains to show that $u_i \cdot w_i \in v^*$ for every $i \in [m]$. 

\begin{claim}\label{claim:cyclicity-bihom-prefix-times-suffix-is-v}
    For every $i \in [m]$, we have $(u_i w_i) \in \Gamma^*$. 
\end{claim}

\begin{proof}
    If $h_{i,i} = \varepsilon$, then $u_i,w_i = \varepsilon$ and the statement holds. Otherwise, pick any $a : A_i$ such that $|h_{i,i}(a,a)| > 0$. We have $h_{i,i}(a,a) = \overline{u_i} v^k \overline{w_i}$ for some $k \geq 0$. Note that $h^*(a^M) = h_{i,i}(a,a)^{M-1} \leq_\infix v^\omega$ for every $M \geq 2$, so $(\overline{u_i} v^k \overline{w_i})^\omega = v^\omega$. Then by Fine-Wilf's Theorem, $(\overline{u_i} v^k \overline{w_i})$ is a power of a cyclic shift of $v$. In particular, $|u_i| + |w_j| \equiv 0 \, (\mathrm{mod} \, |v|)$. Therefore, since $u_i$ is a prefix of $v^\omega$ and $w_i$ is a suffix of $v^\omega$, the statement follows.
\end{proof}

\subsubsection{Completeness for cyclicity assertions}

In this section, we prove that the axioms we introduced are sufficient for our system to be complete for assertions ``$f$ is $v$-cyclic'' for functions with values in the free monoid.

\begin{lemma}\label{lem:cyclicity-completeness}
    Suppose that $f : A \xrightarrow{\mathrm{sf}} \Gamma^*$ is $v$-cyclic. Then it can be proved in our system.
\end{lemma}

The proof of \cref{lem:cyclicity-completeness} follows a similar structure to the proof of \cref{lem:unary-is-the-same-as-cyclic} presented in the previous section. However, there are some new technical caveats that we will handle with the following simple lemma.

\begin{lemma}\label{lem:unbounded-have-fixed-prefix-making-them-cyclic}
    Suppose that $v \in \Gamma^*$ is root-free, $f : A \xrightarrow{\mathrm{sf}} \Gamma^*$ is unbounded and $u, u' \in \Gamma^*$ are such that $(u f), (u' f) \leq_\prefix v^\omega$. Then $|u| \equiv |u'| \, (\mathrm{mod} \, |v|)$. Similarly if $(f u), (f u') \leq_\suffix v^{-\omega}$, then $|u| \equiv |u'| \, (\mathrm{mod} \, |v|)$.
\end{lemma}

\begin{proof}
    We only prove the first statement, as the second is symmetric. Take $a \in \dom(f)$ such that $f(a)$ contains a copy of $v$. Looking at how $u \cdot f(a)$ aligns with $v^\omega$, we see that the copy of $v$ in $f(a)$ must start at an index divisible by $|v|$ --- otherwise we would have that $v$ is a proper infix if $v^2$, contradicting the fact that it is root-free by \cref{lem:prime-words-must-perfectly-align}. Therefore, $|u| \, (\mathrm{mod} \, |v|)$ depends only on $f(a)$, and the same is true for $|u'|$.
\end{proof}

\begin{proof}[Proof of \cref{lem:cyclicity-completeness}]
    The proof follows a nested induction on the rank of $A$, and then the number of subfunctions in $f$. However, we need to slightly strenghten the inductive hypothesis: we will prove that if $(u f w)$ is $v$-cyclic for some $u, w \in \Gamma^*$ with $|u| \in \{0, \dots, |v|-1\}$ and $|w| \in \{1, \dots, |v| \}$, then our system can prove it. We start with the induction basis: if $A$ is an atomic type, or more generally, $f$ is a constant function, then the statement follows directly from axioms. Next, we consider concatenations and bihomomorphisms separately.

    \paragraph{Concatenation.} Suppose that $f = (g h)$. If both $g,h$ are constant, then so is $f$, and the statement follows. Otherwise, we know that $g$ and $h$ are $v$-unary. By \cref{lem:unary-is-the-same-as-cyclic}, there are $u_0,u_1,w_0,w_1$ such that $(u_0 g w_0)$ and $(u_1 h w_1)$ are $v$-cyclic, both of which can be proved in our system by induction. Assume that $|u_0|, |u_1| \in \{0, \dots, |v|-1\}$ and $|w_0|, |w_1| \in \{1, \dots, |v| \}$. 

    Suppose that one of $g,h$ is constant, while the other is unbounded. We only consider the case where $g$ is unbounded and $h$ is constant. Suppose that $g = c$ for some $c \in \Gamma^*$. Since $g$ is unbounded and $(u g), (u_0 g) \leq_\prefix v^\omega$, by \cref{lem:unbounded-have-fixed-prefix-making-them-cyclic} we deduce that $|u| \equiv |u_0| \, (\mathrm{mod} \, |v|)$, and therefore $u = u_0$. By \cref{lem:unarity-you-can-attach-prefix-to-unbounded}, $(u g c)$ is $v$-unary and by \cref{claim:cyclicity-concat-attaching-constant}, there is $w_2 \in \Gamma^*$ such that $(u g (c w_2))$ is $v$-cyclic. We can assume that $|w_2| \in \{1, \dots, |v| \}$. We have $(g w_0), (g (c w_2)) \leq_\suffix v^{-\omega}$, so by \cref{lem:unbounded-have-fixed-prefix-making-them-cyclic} we have $w_0 \cdot v^k = c w_2$ for some $k \geq 0$. Therefore, the fact that $(u g (c w_2))$ is $v$-cyclic is a consequence of \nameref{rule:unarity-concat} for $(u g w_0)$ and $v^k$. However, $(u g (c w_2)) = (u f w_2) = (u f w)$, because $w = w_2$ by \cref{lem:unbounded-have-fixed-prefix-making-them-cyclic}. 

    Finally, it remains to consider the case that both $g,h$ are unbounded. First of all, by \cref{lem:unbounded-have-fixed-prefix-making-them-cyclic} we have $u = u_0$ and $w = w_1$. Furthermore, by \cref{lem:unarity-you-can-attach-prefix-to-unbounded} we have that $(u g \cdot h w)$ is $v$-unary, and by \cref{claim:cyclicity-concant-both-unbounded} applied to $f := (u g), g := (h w)$ we deduce that $(u_1 w_0) = v$. The rest follows from \nameref{rule:unarity-concat}. 

    \paragraph{Bihomomorphisms.} Suppose that $f = \bihom{to, h} : A_1 \times (A_1 + \dots + A_m)^* \times A_1 \to \Gamma^*$ (the case of reversed bihomomorphism is symmetric). We can assume that $f \neq \varepsilon$, and therefore $f$ is unbounded. The function $h$ can be viewed as a collection of simple functions $h_{i,j} : A_i \times A_j \to \Gamma^*$.

    By \cref{claim:cyclicity-bihom-prefix-and-suffix-depends-only-on-bound-type}, there are $u_1, \dots, u_m, w_1, \dots, w_m$ such that $(u_i \cdot h_{i,j} \cdot w_j)$ is $v$-cyclic for every $i,j \in [m]$. All of these assertions can be proved by induction. We can assume that $|u_i| \in \{0, \dots, |v|-1 \}$ and $|w_i| \in \{1, \dots, |v| \}$ for every $i \in [m]$. By \cref{claim:cyclicity-bihom-prefix-times-suffix-is-v}, we get that $(u_i w_i) = v$ for every $i \in [m]$. Therefore, $(u_1 f w_1)$ is $v$-cyclic and this follows from \nameref{rule:unarity-bihom}. However, by \cref{lem:unbounded-have-fixed-prefix-making-them-cyclic} we obtain $u = u_1, w = w_1$, which finishes the proof.
\end{proof}

\begin{comment}
\subsubsection{System for proving cyclicity}

First, we give an equivalent definition of unarity in terms of functions equality.

\begin{lemma}\label{lem:unarity-in-terms-of-functions-equality}
    $\im \left(u \cdot f \cdot w \right) \subseteq v^*$ if and only if $v^{|u f w|} = (u \cdot f \cdot w)^{|v|}$.
\end{lemma}

\begin{proof}
    The length of both sides is equal: it is $|v| \cdot |u \cdot f \cdot w|$. This suffices for $\implies$. Now suppose that the right hand side holds. Then $v^\omega = (u \cdot f \cdot w)^\omega$, and therefore by the Fine-Wilf's Theorem, words $v$ and $u \cdot f \cdot w$ have a common root of length $\mathrm{gcd}(|v|, |u \cdot f \cdot w|)$. By the fact that $u$ is primal, this root must have length $|v|$, and hence $u \cdot f \cdot w \in v^*$.
\end{proof}

Thanks to \cref{lem:unarity-in-terms-of-functions-equality}, we can attempt to prove unarity in our equational system. Everytime we write $\im(u \cdot f \cdot w) \subseteq v^*$, we treat it as sugar syntax for the equational definition.

Now, we introduce axioms that we need to add to our system to prove unarity.

\[
    \im (v) \subseteq v^{(*)} 
\]

\[
    \im(\overline v) \subseteq v^{(*)}
\]

\begin{mathpar}
    \inferrule
        {   
            \im (f) \subseteq v^{(*)} \\ 
            \im (g) \subseteq v^{(*)} 
        }
        {
            f \cdot g = g \cdot f
        }[\textsc{(unarity-comm)}]\label{rule:unarity-commutativity}

    \inferrule
        {   
            \im (h) \subseteq v^{(*)}
        }
        {
            \im \left( \bihom{to, h} \right) \subseteq v^{(*)}
        }[\textsc{(unarity-bihom-lift)}]\label{rule:unarity-bihom-lift}

    \inferrule
        {   
            \im (g) \subseteq v^{(*)} \\ 
            \im (h) \subseteq v^{(*)}
        }
        {
            \bihom{to, g \cdot h} = \bihom{to, g} \cdot \bihom{to, h}
        }[\textsc{(unarity-bihom-dist)}]\label{rule:unarity-bihom-dist}
\end{mathpar}

We introduce two lemmas used for proving unarity in our system and prove that they follow from our axioms.

\begin{mathpar}
    \inferrule
        {
            \im \left( u \cdot f \cdot x \right) \subseteq v^* \\\\ 
            \im \left( y \cdot g \cdot w \right) \subseteq v^* \\\\ 
            y \cdot x = v
        }
        {
            \im \left(u \cdot fg \cdot w \right) \subseteq v^*
        }[\textsc{(unarity-concat)}]\label{rule:unarity-concat}

        \inferrule
        {
            \forall i,j \quad \im \left( u_i \cdot h_{| A_i \times A_j} \cdot w_j \right) \subseteq v^* \\\\
            \forall i \quad u_i \cdot w_i = v
        }
        {
            \im \left( u_1 \cdot \bihom{to, h} \cdot w_1 \right) \subseteq v^*
        }[\textsc{(unarity-bihom)}]\label{rule:unarity-bihom}
\end{mathpar}

\begin{lemma}\label{lem:unarity-concat-follows-from-axioms}
    \nameref{rule:unarity-concat} follows from axioms.
\end{lemma}

\begin{proof}
    \begin{align*}
        \left( u \cdot f g \cdot w \right)^{|v|} &= \left( u \cdot f \cdot x \cdot \overline{xy} \cdot y \cdot g \cdot w \right)^{|v|} \\ 
                                                     &= \left( \left( u \cdot f \cdot x \right) \cdot \overline v \cdot \left( y \cdot g \cdot w \right) \right)^{|v|} \tag{by $y \cdot x = v$} \\ 
                                                     &= \left( u \cdot f \cdot x \right)^{|v|} \cdot \overline{v}^{|v|} \cdot \left( y \cdot g \cdot w \right)^{|v|} \tag{by \nameref{rule:unarity-commutativity}} \\ 
                                                     &= v^{|u f x|} \cdot v^{-|v|} \cdot u^{|y g w|} \tag{by assumptions} \\ 
                                                     &= u^{|u fg w|} \tag{by $|x| + |y| = |v|$}
    \end{align*}
\end{proof}

\begin{lemma}\label{lem:unarity-bihom-follows-from-axioms}
    \nameref{rule:unarity-bihom} follows from axioms.
\end{lemma}

\begin{proof}
    \begin{align*}  
        \left( u_1 \cdot \bihom{to, h} \cdot w_1 \right)^{|v|} &= \left( u_1 \cdot w_1 \cdot \bihom{to, \overline{w_i} \cdot h_{| A_i \times A_j} \cdot w_j } \right)^{|v|}  \tag{by \eqref{rule:bihom-rotation}} \\ 
                                          &= \left( v \cdot \bihom{to, \overline v \cdot u_i \cdot h_{| A_i \times A_j} \cdot w_j } \tag{by $u_i \cdot w_i = v$} \right)^{|v|} \\
                                          &= \left( v \cdot \bihom{to, \overline v} \cdot \bihom{to, u_i \cdot h_{| A_i \times A_j} \cdot w_j } \right)^{|v|} \tag{by \nameref{rule:unarity-bihom-dist}} \\ 
                                          &= v^{|v|} \cdot \bihom{to, \overline{v}}^{|v|} \cdot \bihom{to, u_i \cdot h_{| A_i \times A_j} \cdot w_j }^{|v|} \tag{by \nameref{rule:unarity-commutativity}} \\
                                          &= v^{|v|} \cdot \bihom{to, \overline{v}^{|v|}} \cdot \bihom{to, \left(u_i \cdot h_{| A_i \times A_j} \cdot w_j\right)^{|v|} } \tag{by \nameref{rule:unarity-bihom-dist}} \\
                                          &= v^{|v|} \cdot \bihom{to, v^{-|v|}} \cdot \bihom{to, v^{|u_i| + |h_{|A_i \times A_j}| + |w_j|}} \tag{by assumptions} \\
                                          &= v^{|v|} \cdot \bihom{to, v^{|h_{|A_i \times A_j}|}} \cdot \bihom{to, v^{|u_i| + |w_j| - |v|}} \tag{by \nameref{rule:unarity-bihom-dist} and \nameref{rule:unarity-commutativity}} \\
                                          &= v^{|v|} \cdot v^{|\bihom{to, h}|} \cdot v^{|u_1|} \cdot \bihom{to, v^{|u_j| + |w_j| - |v|}} \cdot v^{-|u_1|} \tag{by \eqref{rule:bihom-rotation}} \\ 
                                          &= v^{|v| + |\bihom{to, h}|} \tag{by $|u_j| + |w_j| = |v|$} \\ 
                                          &= v^{|u_1 \cdot \bihom{to, h} \cdot w_1|} \tag{by $|u_1| + |w_1| = |v|$}
    \end{align*}
\end{proof}

Finally, we prove completeness, in some even stronger version. We show that in fact, $\im(f) \subseteq \infix(u^\omega)$ suffices to be unary.

\begin{lemma}\label{lem:unarity-completeness}
    If $\im(f) \subseteq \infix \left(v^\omega \right)$, then $\im \left(u \cdot f \cdot w \right) \subseteq v^*$ for some $u, w \in \Gamma^*$ and it can be derived from our axioms.
\end{lemma}

\begin{proof}
    The proof follows an induction of the rank of type. If $f$ is a constant function, then trivial. Otherwise we consider concatenation and bihomomorphism separately.

    \paragraph{Concatenation.} Suppose that $\im(f \cdot g) \subseteq \infix \left( v^\omega\right)$. Then $\im(f), \im(g) \subseteq \infix \left( v^\omega\right)$, so from induction we have $u_0, w_0, u_1, w_1$ such that $\im(u_0 \cdot f \cdot w_0) \subseteq v^*$ and $\im(u_1 \cdot g \cdot w_1) \subseteq v^*$. We can assume that $|w_0| + |u_1| \in \{ 1, |v|+1, \dots, 2|v| - 1\}$. If both $f,g$ are bounded, then they are constant and the statement follows.
    
    Suppose that one of $f,g$ is bounded. Assume that $g$ is bounded (the other case is symmetric). Then $g$ is constant, equal to some $c \in \Gamma^*$. By considering an input such that $f$ contains a copy of $v$, we obtain that $c$ is in prefix relation with $w_0$. If $c \leq_\prefix w_0$, then $\im(u_0 \cdot f \cdot c \cdot (\overline{c} \cdot w_0)) \subseteq v^*$ and it can be proved by induction. If $w_0 <_\prefix c$, then the same argument works if we take $w_0 = w_0 \cdot v^k$ for some constant $k$ (to make it longer than $c$).

    Now suppose that both $f$ and $g$ are unbounded. By \nameref{rule:unarity-concat}, it suffices to show that $|v| \, | \, |w_0| + |u_1|$. To see that this holds, we consider an input such that both $f$ and $g$ have a copy of $v$ in them. Then we argue that since there is only one way for $f$ and $g$ to map into $v^\omega$ and $(f \cdot g), (f \cdot w_0 \cdot u_1 \cdot g) \leq_\infix u^\omega$, it must be the case that $|v| \, | \, |w_0 \cdot u_1|$.

    \paragraph{Bihomomorphism.} Suppose that $\im(\bihom{to, h}) \subseteq \infix \left( v^\omega \right)$ (the case that bihomomorphism is reversed is symmetric). By induction, for every $i,j$, there are $u_{i,j},w_{i,j}$ such that $\im(u_{i,j} \cdot h_{|A_i \times A_j} \cdot w_{i,j}) \subseteq v^*$. We can assume that $|u_{i,j}| \in \{1, \dots, |v| \}$ and $|w_{i,j}| \in \{0, \dots, |v|-1 \}$. \katzpermargin{TODO: Finish}
\end{proof}

\begin{comment}
\begin{lemma}\label{lem:weak-unarity-implies-strong-unarity}
    Suppose that $f : A \xrightarrow{\mathrm{sf}} \Gamma^*$ and $u \in \Gamma^*$ are such that $\im(f) \subseteq_{\infix} u^\omega$. Then there is a cyclic shift $v$ of $u$ and $v_0 <_\suffix v$, $v_1 <_\prefix v$ such that
    \[
        \im(f) \subseteq v_0 \cdot v^* \cdot v_1.
    \]
\end{lemma}

\begin{proof}
    We proceed by structural induction, but we also want to consider types in order of their rank. If $f$ is a constant, then the statement is trivial. Otherwise, we separately consider concatenations and bihomomorphisms.

    \paragraph{Concatenation.} We prove the statement for $f \cdot g$, where $f,g : A \xrightarrow{\mathrm{sf}} \Gamma^*$. By induction, there are cyclic shifts $v,w$ of $u$ and $v_0 <_\suffix v, v_1 <_\prefix v, w_0 <_\suffix w, w_1 <_\prefix w$ such that $\im(f) \subseteq v_0 \cdot v^* \cdot v_1$ and $\im(g) \subseteq w_0 \cdot w^* \cdot w_1$. 
    
    If both $f,g$ are bounded, then they are constant, and the statement follows. Assume that one of the functions is constant, while the other is unbounded. Assume that $f$ is unbounded (the other case is symmetric) and $g = c$. By considering an input $a : A$ such that $|f(a)| \geq 2|v|-1$, we argue that $v_1 \cdot c <_\prefix v^*$. Therefore,
    \[
        \im(f \cdot g) \subseteq v_0 \cdot v^* \cdot v_1 \cdot c \subseteq v_0 \cdot v^* \cdot \strpref{v}{|v_1| + |c| \, (\mathrm{mod} \, |v|)}.
    \]
    It remains to consider the case where both $f$ and $g$ are unbounded. Suppose that $w$ is the $i$-th cyclic shift of $v$, i.e. that $w = \strsuf{v}{i+1} \cdot \strpref{v}{i}$. \katzpermargin{is it easy to verify that if $v,w$ are prime, then $i$ is unique} We take $a  :A$ such that $|f(a)|, |g(a)| \geq 2|v|-1$. Consider $(f \cdot g)(a)$ --- between the last occurrence of $v$ in $f(a)$ and the first occurrence of $w$ in $g(a)$, we have exactly $v_1 \cdot w_0$. Therefore, we have $v_1 \cdot w_0 \in v^* \cdot \strpref{v}{i}$. Moreover, we have $w_1 <_\prefix \strsuf{v}{i+1} \cdot v^*$. Thus,
    \begin{align*}
        \im(f \cdot g) &\subseteq v_0 \cdot v^* \cdot (v_1 \cdot w_0) \cdot w^* \cdot w_1 \\ 
                       &\subseteq v_0 \cdot v^* \cdot \strpref{v}{i} \cdot w^* \cdot w_1 \\ 
                       &\subseteq v_0 \cdot v^* \cdot \strpref{v}{i} \cdot w_1 \\ 
                       &\subseteq v_0 \cdot v^* \cdot \strpref{v}{i + |w_1| \, (\mathrm{mod} \, |v|)}.
    \end{align*}    

    \paragraph{Bihomomorphism.} Suppose that $A = A_1 \times (A_1 + \dots + A_m)^* \times A_1$ and $f$ is a bihomomorphism. If $f = \varepsilon$, the statement holds. Otherwise, we choose any $c = (a_0 \cdot b \cdot a_0) : A$ such that $|f(c)| \geq 2|u|-1$. Because of the length and $f(c) <_\infix u^\omega$, we have $u \leq_\infix f(c)$. Since $u$ is prime, there are unique $u_0 <_\suffix u$ and $u_1 <_\prefix u$ with $f(c) \in u_0 \cdot u^+ \cdot u_1$. 
    Now consider: 
    \[
        u^\omega >_\infix f(a_0 \cdot b \cdot a_0 \cdot b \cdot a_0) = f(c)^2 \in u_0 \cdot u^+ \cdot (u_1 \cdot u_0) \cdot u^+ \cdot u_1,
    \]
    we obtain that $u_1 \cdot u_0 \in \{\varepsilon, u\}$. In both cases, we get that $f(c) \in v^*$ for some cyclic shift $v$ of $u$.

    We now argue that $\im(f) \subseteq v^*$. Pick any $a : A$. Suppose that $a = (a_1, \dots, a_n)$ and consider a function $g : \prod_{i=1}^n \type{a_i} \to \Gamma^*$ which agrees with $f$ on inputs of length exactly $n$ with types of elements exactly the same as in $a$. We have $\im(g) \subseteq_\infix v^\omega$ and $g$ is defined on a type of smaller rank, so from induction there are $w, w_0, w_1$ such that $\im(g) \subseteq w_0 \cdot w^* \cdot w_1$ and $w$ is a cyclic shift of $v$. In particular, we have $f(a) \in w_0 \cdot w^* \cdot w_1$.

    By exchanging $\first{a}$ and $\last{a}$ to $a_0$, we obtain an input $a' \in \dom(g)$. So $f(a') \in w_0 \cdot w^* \cdot w_1$. By arguments similar as previously, we show that $w_1 \cdot w_0 \in \{\varepsilon, w\}$. Moreover, we consider 
    \[
        v^\omega >_\infix f(a_0 \cdot b \cdot a' \cdot b \cdot a_0) = f(c) \cdot f(a) \cdot f(c) \in v^+ \cdot f(a) \cdot v^+,
    \]
    so we have $f(a) \in v^*$.

    If $w_1 \cdot w_0 = \varepsilon$, then $f(a') \in v^* \cap w^*$, so $f(a') = \varepsilon$ or $w = v$. The latter gives $f(a) \in v^*$.

    If $w_1 \cdot w_0 = w$, then $f(a') \in (w_0 \cdot w_1)^* \cap v^*$, so $f(a') = \varepsilon$ or $w_0 \cdot w_1 = v$. The latter gives $w_0 \cdot w^* \cdot w_1 = v^+ \implies f(a) \in v^*$.

    It remains to solve the case where $f(a) = \varepsilon$. This case can be easily avoided if $f(A_1 \times A_1) \neq \varepsilon$ --- by chosing an appropriate $a_0$ at the very beginning of the proof. However, if $f(A_1) = \varepsilon$, then we have $f(a') = f(a)$, but we already established that $f(a) \in v^*$. This finishes the proof.
\end{proof}
\end{comment}